### 2018-03-14 paper/3-compute.tex
Work on S3

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in the previous
section~\S\ref{sec-background} and design complex build systems from simple
primitives. Specifically, we present the \emph{compute} and \emph{build}
abstractions, which are described in
Sections~\ref{sec-task} and \ref{sec-general-build} respectively.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}

\hs{Build} corresponds to the algorithm used for bringing a key-value store
\store to a consistent state by recomputing out-of-date values in an
appropriate order. We explain the build type and use it to develop a formal
definition of build systems in~\S\ref{sec-general-build}.
Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the build
abstraction further and provide concrete implementations for several build
systems.

\subsection{Common vocabulary for build systems}
\label{sec-vocabulary}

\begin{figure}
\begin{minted}{haskell}
-- Stores
data Store i k v   -- Abstract
getInfo  :: Store i k v -> i
putInfo  :: Store i k v -> i -> Store i k v
getValue :: Store i k v -> k -> v
putValue :: Hashable v => Store i k v -> k -> v -> Store i k v
getHash  :: Hashable v => Store i k v -> k -> Hash

-- Hashing
hash :: Hashable a => a -> Hash

-- Applicative
(<*>) :: Applicative f => f (a->b) -> f a -> f b  -- Left-associative
(<$>) :: Applicative f => (a->b) -> f a -> f b    -- Left-associative
pure  :: Applicative f => a -> f a

-- State monad
data State s
instance Applicative State
instance Monad       State
evalState :: State s a -> s -> s
getState  :: State s s
\end{minted}
\caption{Signatures of library functions}
\label{fig-lib-funs}
\end{figure}

We start by establishing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets.
% We will use this vocabulary throughout the rest of the paper.

A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up to date.  The store is a key/value mapping, togther with
\emph{persistent build state} that is maintained by the build system itself.

In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.

@@

The types of the library functions we use in this paper is
given in Figure~\ref{fig-lib-funs}.
We use a cryptographic \emph{hash function} \hs{hash :: Hashable v =>
v -> Hash} for efficient tracking and sharing of~build results, where
\hs{Hash} is an abstract data type equipped with the equality test.

A \emph{store}, of type \hs{Store i k v}, associates keys to values.
It is convenient to assume that a
store is total function \store, i.e. it contains a value for every possible key.
We therefore also assume that the type of values is capable of encoding values
corresponding to non-existent keys (missing files or empty cells). This, in
particular, suggests that it is possible to \emph{depend on the absence of a
value}, which is a useful feature for build systems. In addition to usual
read and write operations, some stores also support a dedicated \emph{read hash}
operation, which is implemented more efficiently than by reading and hashing the
result. For example, GVFS (Git Virtual File System)~\cite{gvfs} stores file
hashes and downloads file contents only on demand -- an important feature for
cloud build systems.

The store also contains a single persistent value, of type \hs{i}, representing
the persistent state maintained by the build system itself -- its ``memory''.

@@

Our first key idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type is best introduced through an example.
Consider this Excel spreadsheet (c.f. \S\ref{sec-background-excel}):
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{bash}
  A1: 10     B1: =A1 + A2
  A2: 20     B2: =B1 * 2
\end{minted}
\vspace{1mm}
Here cell \hs{A1} contains the value \hs{10} while cell \hs{B1} contains
the formula \hs{=A1+A2}.
We can represent the formulae of of this spreadsheet
with the following task description
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Compute Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _   _      = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the constant values.  Like every \hs{Task}, \hs{sprsh1} is given
a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
a task description (in the Excel case, a formula)
for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
that cell.  If there is a formula in the cell, it computes
the value of the formula, using \hs{fetch} to find the value of any keys on
which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures
are given in Figure~\ref{fig-lib-funs}.  We will return to the need for this
generality in
Section~\ref{sec-deps}\footnote{Readers familiar with \emph{lenses} or \emph{profunctor optics}
might recognise a familiar pattern. We discuss this in~\S\ref{sec-related-optics}.}.

For now, we content ourselves with observing that a task description,
of type \hs{Task c k v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a call-back (\hs{fetch}) to obtain the values
of its dependencies.

@@

Next comes our second key abstraction: a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up-to-date.  (What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\emph{SLPJ: is this right?}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
dumbBuild :: Build Monad () k v
dumbBuild task k store = execState (go k) store
  where
    go :: k -> State s v
    go k = case (task go k) of
             Nothing  -> do { s <- getState; return (getValue s k) }
             Just act -> do { v <- act; updState (\s -> putStore s k v); return v }
\end{minted}
\vspace{1mm}
\emph{SLPJ: explain}

@@

\subsection{Computing dependencies}

>>>>

This section presents purely functional abstractions that allow us to express
all the intricacies of build systems discussed in the previous
section~\S\ref{sec-background} and design complex build systems from simple
primitives. Specifically, we present the \emph{task} and \emph{build}
abstractions in~\S\ref{sec-task} and~\S\ref{sec-general-build}, respectively.
Sections~\S\ref{sec-build} and~\S\ref{sec-implementations} scrutinise the
abstractions further and provide concrete implementations for several build
systems.

We start by establishing a common vocabulary for build systems that allows
us to abstract away from specific application domains, such as software
development or spreadsheets.

\subsection{Common vocabulary for build systems}\label{sec-vocabulary}

\begin{figure}
\begin{minted}{haskell}
-- An abstract store
data Store i k v
getInfo    :: Store i k v -> i
putInfo    :: Store i k v -> i -> Store i k v
getValue   :: Store i k v -> k -> v
putValue   :: (Eq k, Hashable v) => Store i k v -> k -> v -> Store i k v
getHash    :: Hashable v => Store i k v -> k -> Hash
initialise :: Hashable v => i -> (k -> v) -> Store i k v
\end{minted}
\vspace{1mm}
\begin{minted}{haskell}
-- Hashing
hash :: Hashable a => a -> Hash
\end{minted}
\vspace{1mm}
\begin{minted}{haskell}
-- Applicative functors
pure  :: Applicative f => a -> f a
(<$>) :: Applicative f =>   (a -> b) -> f a -> f b -- Left-associative
(<*>) :: Applicative f => f (a -> b) -> f a -> f b -- Left-associative
\end{minted}
\vspace{1mm}
\begin{minted}{haskell}
-- State monad
data State s
instance Applicative (State s)
instance Monad       (State s)
get       :: State s s
put       :: s -> State s ()
modify    :: (s -> s) -> State s ()
execState :: State s a -> s -> s
\end{minted}
\vspace{-2mm}
\caption{Signatures of library functions}\label{fig-lib-funs}
\vspace{-4mm}
\end{figure}

A \emph{build system} takes a \emph{task description}, a target \emph{key},
and a \emph{store}, and returns a new store in which the target key is
up to date. The store is a key-value map, together with \emph{persistent build
information} that is maintained by the build system itself.

In software build systems, keys are typically filenames,
e.g.~\cmd{main.c}, and values are file contents (a C program source
code in this case). In spreadsheets, keys are cell names,
e.g. \cmd{A1}, and values are numbers, text, etc. that are typically
displayed inside cells. In our Haskell code, we will use type
variables \hs{k} and \hs{v} to denote keys and values, respectively.
Fig.~\ref{fig-lib-funs} provides a Haskell encoding of our main definitions,
and gives types of the library functions we use in this paper.

@@

We use a cryptographic \emph{hash function}
\hs{hash}~\hs{::}~\hs{Hashable}~\hs{v}~\hs{=>}~\hs{v}~\hs{->}~\hs{Hash} for
efficient tracking and sharing of build results, where \hs{Hash} is an abstract
data type equipped with the equality test.

A \emph{store}, of type \hs{Store}~\hs{i}~\hs{k}~\hs{v}, associates keys to
values. It is convenient to assume that a store is total, i.e. it contains a
value for every possible key. We therefore also assume that the type of values
is capable of encoding values corresponding to non-existent keys (missing files
or empty cells). This, in particular, suggests that it is possible to
\emph{depend on the absence of a value}, which is a useful feature for build
systems. In addition to usual \hs{getValue} and \hs{putValue} operations, some
stores also support a dedicated \hs{getHash} operation, which is implemented
more efficiently than by hashing the result of \hs{getValue}. For example, GVFS
(Git Virtual File System)~\cite{gvfs} stores file hashes and downloads file
contents only on demand. The store also contains a single persistent value, of
type \hs{i}, representing the persistent state maintained by the build system
itself~--~its~``memory''. We can \hs{initialise} a store by supplying an initial
value \hs{i} and a key-value map \store.

@@

Our first main idea is the task-description abstraction:
\begin{minted}[xleftmargin=10pt]{haskell}
type Task c k v = @\std{forall}@ f. c f => (k -> f v) -> k -> Maybe (f v)
\end{minted}
This highly-abstracted type\footnote{Readers familiar with \emph{lenses} or
\emph{profunctor optics} might recognise a familiar pattern. We discuss this
in~\S\ref{sec-related-optics}.} is best introduced by an example.
Consider this \Excel spreadsheet:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{text}
  A1: 10     B1: A1 + A2
  A2: 20     B2: B1 * 2
\end{minted}
\vspace{1mm}
Here cell \cmd{A1} contains the value \cmd{10}, the cell \cmd{B1} contains
the formula \cmd{A1+A2}, etc. We can represent the formulae of this spreadsheet
with the following task description:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
sprsh1 :: Compute Applicative String Integer
sprsh1 fetch "B1" = Just ((+)   <$> fetch "A1" <*> "A2")
sprsh1 fetch "B2" = Just ((* 2) <$> fetch "B1")
sprsh1 _   _      = Nothing
\end{minted}
\vspace{1mm}
Here we instantiate keys \hs{k} with \hs{String}, and values \hs{v} with \hs{Integer}.
(Real spreadsheet cells would contain a wider range of values, of course.)
The task description \hs{sprsh1} embodies all the \emph{formulae} of the spreadsheet,
but not the input values.  Like every \hs{Task}, \hs{sprsh1} is given
a \hs{fetch} function and a key. It pattern-matches on the key to see if it has
a task description (in the \Excel case, a formula)
for it.  If not, it returns \hs{Nothing}, indicating that there is no formula in
that cell.  If there is a formula in the cell, it computes
the value of the formula, using \hs{fetch} to find the value of any keys on
which it depends.

The code to ``compute the value of a formula'' in \hs{sprsh1} looks a bit mysterious
because it takes place in an \hs{Applicative} computation \cite{mcbride2008applicative}
-- the relevant type signatures
are given in Fig.~\ref{fig-lib-funs}. We will return to the need for this
generality in~\S\ref{sec-deps}.

For now, we content ourselves with observing that a task description,
of type \hs{Task c k v}, is completely isolated from the world of
compilers, calculation chains, file systems, caches, and all other
complexities of real build systems.  It just computes a single output, in
a side-effect-free way, using a callback (\hs{fetch}) to find the values
of its dependencies.

@@

Next comes our second main abstraction -- a build system:
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
type Build c i k v = Task c k v -> k -> Store i k v -> Store i k v
\end{minted}
\vspace{1mm}
The signature is very straightforward.  Given a task description, a target key,
and a store, the build system returns a new store in which the value of the
target key is up-to-date. What exactly does ``up to date'' mean?  We answer
that precisely in \S\ref{sec-build-correctness}.

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
busy :: (Eq k, Hashable v) => Build Monad () k v
busy task key store = execState (go key) store
  where
    go k = case task go k of
        Nothing  -> do { s <- get; return (getValue s k) }
        Just act -> do { v <- act; modify (\s -> putValue s k v); return v }
\end{minted}
\vspace{1mm}

\emph{SLPJ: explain}

@@

\subsection{Computing dependencies}\label{sec-deps}
